---
title: "Heart Disease Indicators by Sex"
output: github_document
categories: [EDA, Healthcare, R, Decision Trees]
date: "2024-07-18"
---



## Introduction
Heart disease is a leading cause of mortality worldwide, yet significant gender disparities exist in its study and treatment. Historically, research has predominantly focused on male patients, resulting in diagnostic and treatment protocols that do not fully address the unique ways in which heart disease manifests in women. This oversight has led to a critical gap in our understanding and care of heart disease in women.


The original data can me found [here](https://archive.ics.uci.edu/dataset/45/heart+disease). I have already saved it as a csv file and will be accessing this file. Mutliple researchers have explored this data and gotten good predictions on the data! 

__The goal of this project is not to maximize prediction accuracy, but to explore how sex plays a role in heart disease indicators.__ By doing so, we aim to provide medical professionals with deeper insights into the early indicators of heart disease and ensure that sex-specific factors are considered for improved diagnosis and treatment.

To get started lets read in the data and take a look at the first 5 rows.

```{r message=FALSE, warning=FALSE, include=TRUE}
packages <- c("tidyr", "readxl", "dplyr", "magrittr", "purrr", 
              "ggplot2", "stringr", "snakecase", "knitr") 
invisible(lapply(packages, require, character.only = TRUE ))

```


```{r }
df_raw <- read.csv("data/heart_disease.csv")
kable(df_raw[1:5,], caption = "Raw Data")

na_cols <- names(which(colSums(is.na(df_raw)) > 0))
print(na_cols)
```
The data consists of 14 columns, including `num` which is our heart disease number. Two of these columns have `NA` values: `ca` and `thal`. Lets clean up the data and start exploring what we have.

## Data Cleaning and Initial Observations
First, let's label the data in an easier way to understand for `num` (heart disease), and `sex`.

```{r warning=FALSE}
# Add yes/no heart disease column
# Label sex as male/female for convenience
df <- df_raw %>% 
  mutate( heart_disease = case_when(
    num == 0 ~"Present", T ~"None"),
    sex = case_when(sex==1 ~"male", T ~"female"))

# View data with NA values
kable(df %>% filter(if_any(everything(), is.na)))
# Remove NAs
df %<>% drop_na()

# Plot heart disease vs. no disease 
ggplot(df, aes(x = heart_disease, fill = sex))+
  geom_bar(position = position_dodge())+
  ggtitle("Count of Detected Heart Disease for Women and Men")+xlab("Heart Disease")

# Heart Disease numbers by sex
men <- paste0("Men with HD:  ", df %>% filter(sex == 'male', heart_disease == 'Present') %>% nrow())
women <- paste0("Women with HD:  ", df %>% filter(sex == 'female', heart_disease == 'Present') %>% nrow())
stats_to_print <- paste(men, women, sep = "\n")

cat(stats_to_print[1])
```

```{r }
normalize <- function(x) {
  return ((x - min(x)) / (max(x) - min(x)))
}

# Plot avg across all features for men and women
df %>% select(-num) %>% mutate_at(vars(-sex, -heart_disease), normalize)%>% 
  group_by(sex, heart_disease) %>%  summarise_all(mean) %>%  ungroup() %>% 
  pivot_longer(cols= !c("sex", "heart_disease"), names_to = 'feature', values_to = 'mean') %>% 
  ggplot()+
  geom_point(aes(x = feature, y = mean, color = sex))+coord_flip()+facet_wrap(.~heart_disease)+ xlab("")+ylab("Normalized Mean")+
  ggtitle("Feature Means Split By Heart Disease Present vs. None")
```


During the data cleaning process, six data points were removed due to missing information. While imputation methods could be considered to address this issue, for the purpose of this project, I have chosen to omit this step. The primary goal here is to understand the impact of sex on predicting heart disease, rather than achieving the highest possible prediction accuracy.

An initial review of the dataset reveals a significant gender imbalance, with more men represented in the study than women. This is true for both individuals diagnosed with heart disease and those without. This disparity is noteworthy and we should keep it in mind as we explore further as it could potentially lead to a biased understanding of heart disease indicators and outcomes.

The graph of feature means illustrates differences between feature expression for those with and without heart disease. Additionally, certain features, such as `thal`, `oldpeak`, and `exang`, display distinct variations between men and women. This suggests that the same approach to finding heart disease indicators may not be equally effective for both men and women.


## Decision Tree Models
Bringing it back to our original goal, to explore how sex plays a role in heart disease indicators, I am going to look at a series of decision trees.

### Why Decision Trees?
Decision trees may not always be the best models that give the highest predictive accuracy, however I think are a great model to use to help us understand our problem at hand. 

Decision trees are easy to understand and interpret, making the logic behind the predictions clear. This transparency is particularly valuable in the medical field, where it is crucial to explain why certain features are indicative of heart disease. Unlike more complex models such as neural networks, decision trees do not require dimensionality reduction, and they can directly use the original features of the dataset. This simplicity helps maintain the medical terminology and context, making the results more accessible and actionable for healthcare professionals.


### Building the Models
To understand how women and men are represented within the model, lets build three decision trees:

1. __Full Tree:__ This tree will be built using the training set from all the data.
2. __Male Tree:__ This tree will be built using the training set from the data filtered to `sex == 'male'`.
3. __Female Tree:__ This tree will be built using the training set from the data filtered to `sex == 'female'`

All trees will use a 70/30 training/test split, and use the gini index within the tree. 

> I will not be performing any hyper-parameter tuning in this project. Please take a look at a one of my more in-depth project, such as [this one](https://github.com/lilynorthcutt/sierraNevadaAge), if you are interested in this. 

```{r}


```

## Results and Discussion

## Conclusion and Future Work



